Algebraic Groups\vspace{-1ex}

### 2019-10-01 Algebraic Geometry/Algebraic-Geometry.tex
update

% Document-Specific Macros
\DeclareMathOperator{\1}{\mathbbm{1}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Sym}{Sym}

@@

>>>>

% Document-Specific Macros
\DeclareMathOperator{\1}{\mathbbm{1}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\Sym}{Sym}
\DeclareMathOperator{\Spec}{Spec}
\let\O\relax
\newcommand*{\O}{\calO}
\newcommand{\A}[1]{\bbA^{#1}}
\let\P\relax
\newcommand{\P}[1]{\bbP^{#1}}

@@

\section{September 30th, 2019}
One question that was asked: if we have fixed the underlying topological space in a space with functions, must there be a morphism 
between them somehow? Might there instead be a common cover of the two?

\begin{ex}
	Let $k$ be a field with some topology on it such that every point is closed (you could do the discrete topology).
	Let $\tilde\O(U)$ be the continuous funcitons $U\to k$. In other words, these functions are locally constant.

	Locally constant functions behave nicely under restrictions to opens, of course. The other axioms are also great.

	Have we really found an initial object in our category? This would be enough to establish a ``tent'' (as in 
	localization of categories). Try this out and see what happens!
\end{ex}

\subsection{The question of affine space}
Recall the question about whether $\A 1$ is an affine variety. The idea here is that $\phi:X\to k$ is a morphism of 
spaces with functions if and only if it is regular (that is, in $\O_{\A 1}$).

One direction is tautological (a morphism to $\A1$ has a polynomial underlying it), so let $\phi$ be regular. Then to see that $\phi$ is continuous 
can be checked by pulling back all closed sets. The important observation is that $D(f-a)=X\setminus\phi^{-1}(a)$, which is closed (an axiom for spaces with functions).

The last thing to check is where $\phi$ pulls back regular functions to regular functions. This relies on the facts that $\O_X$ is a $k$-algebra and that 
$\phi(x)-b_j$ is regular on $U$ when $b_j\notin U$.

\subsection{Algebra maps}
Notice that since we have a condition that $\O_X(X)$ must be finitely generated as a $k$-algebra, this means that 
\[\Hom(X,Y)=\Hom_k(\O_Y(Y),\O_X(X))=\Hom_k(k[x_1,\dots,x_n]/(f_1,\dots,f_m),O_X(X))\]
and
\[\Hom(X,Y)=\{(\gamma_1,\dots,\gamma_n)\in(\O_X(X))^n:f_j(\gamma_i)=0,\forall j=1,\dots,m\}\]
In other words, we are looking at maps that factor through $Z$:
\begin{center}
	\begin{tikzcd}
		(\gamma_1,\dots,\gamma_n):X\ar[r]\ar[dr] & k^n\\
		& Z=Z(f_i)\ar[u,hook]
	\end{tikzcd}
\end{center}
Now what we want to say is that $Y=Z$. That is, \textit{affine varieties are closed subsets of affine spaces.}

Now this is all good, but the problem is that we had to \textit{choose} a presentation of $\O_Y(Y)$ to get this picture.
of course we want something more canonical! We will see in this class (and in Kempf) that this can be done.

### 2019-10-01 Algebraic Groups/Algebraic-Groups.tex
update

\begin{defn}
	Let $G$ be a linear algebraic group. Then 
	\begin{itemize}
		\item $G$ is a \textbf{vector group} if $G\cong V_a$ for some finite dimensional $V$.
		\item $G$ is a \textbf{split torus} if $G\cong \bbG_m^n$.
		\item $G$ is a \textbf{torus} if there is a field extention $k\to k'$ such that 
		\[G\times_{\Spec k}\Spec k'\cong \bbG^n_{m,k'}\]
	\end{itemize}
\end{defn}
\end{document}

>>>>

\begin{defn}
	Let $G$ be a linear algebraic group. Then 
	\begin{itemize}
		\item $G$ is a \textbf{vector group} if $G\cong V_a$ for some finite dimensional $V$.
		\item $G$ is a \textbf{split torus} if $G\cong \bbG_m^n$.
		\item $G$ is a \textbf{torus} if there is a field extention $k\to k'$ such that 
		\[G\times_{\Spec k}\Spec k'\cong \bbG^n_{m,k'}\]
	\end{itemize}
\end{defn}

\section{September 30th, 2019}
Another example to consider:
\begin{ex}
	Let $G=\operatorname{PGL}_n$, the projective linear group. Recall we want to define this as $\GL_n/k^\ast$ (from group theory). To do this for algebraic groups,
	we define 
	\[\operatorname{PGL}_n=\operatorname{Proj}k[x_{ij}]_{det}:= \Spec (k[x_{ij}]_{det})_0\]

	The geometric picture is difficult since we haven't yet defined quotients, but as a functor we say $\operatorname{PGL}_n$ 
	is $\Aut(\bbP^n)$, the functor that sends $S\mapsto \Aut(\bbP_S^n)$ where $\bbP^n_S=\bbP_k^n\times_{\Spec k} S$.
\end{ex}

\subsection{Non-affine group schemes}
\begin{ex}
	Let $\lambda\ne 0,1$ be an element in $k$. Then we can define the elliptic curve 
	\[E_\lambda=V(y^2z-x(x-z)(x-\lambda z))\subset \bbP^2\]
	Which gives us a double cover over $(0,1)$ and $(\lambda,\infty)$ with singleton fiber (ramified) over $0,1,$ and $\lambda$.

	Then for any $\lambda\ne 0,1$, $E_\lambda$ is a \textbf{projective} group scheme.
\end{ex}
\begin{rmk}
	If you look at the $\bbC$-points, you get $E_\lambda(\bbC)=\Lambda_\lambda$, giving you a torus. Recall (from e.g. complex analysis) that the 
	moduli here is $\operatorname{SL}_2(\bbZ)$ of all elliptic curves.
\end{rmk}

\subsection{Abelian Varieties}
\begin{defn}
	An \textbf{abelian variety over $k$} is asmooth, geometrically connected ($A\times_{\Spec k}\Spec\bar k$ is connected), proper group scheme $A$ over $k$.
\end{defn}
\begin{ex}
	Over $\bbC$, $\bbC^g/\Lambda$ where $\Lambda\cong \bbZ^{2g}\subseteq \bbC^g$ gives us a genus $g$ example.
\end{ex}
\begin{thm}
	Any abelian variety over $k$ is commutative and projective.
\end{thm}
\begin{thm}[Chevalley]
	If $G$ is any group scheme, then the sequence 
	\[1\to H\to G\to A\to 1\]
	is exact, where $H$ is a linear algebraic group (affine!) and $A$ is an abelian variety.
\end{thm}
\begin{ex}
	Let $X\to \Spec k$ be a geometrically integral projective scheme (proper may suffice). The idea here is that over $\bbC$
	the rings over every open set are integral domains.

	Now consider the \textbf{Picard functor} $\operatorname{Pic}_X:\operatorname{Pic}:\mathbf{Sch/k}\to \Grp$ sending 
	\[S\mapsto \operatorname{Pic}(X_S=X\times_k S)/p^k\operatorname{Pic(S)}\]
\end{ex}
\begin{thm}
	$\operatorname{Pic}_X$ is represented by a scheme locally of finite type, thus $\operatorname{Pic}_X^0$, the connected 
	component of the identity in $[\calO_X]\in\operatorname{Pic}_X$ is an abelian variety.
\end{thm}
\subsection{Relative Group Schemes}
\begin{ex}
	Consider $\bbG_{m,\bbZ}=\Spec \bbZ[t]_t$. Then $G_{m,S}=\bbG_{m,\bbZ}\times_{\Spec \bbZ}S$. 
	In the case that $S=\Spec R$, $\bbG_{m,S}=\Spec R[t]_t$.
\end{ex}

\begin{ex}
	Let $\bbA^1=\Spec k[x]$ and define $G=\Spec k[x,y]_{xy+1}\subseteq\bbA^2$. Notice this is the plane minus a hyperbola.

	Define $\cdot:G\times_{\bbA^1}G\to G$ to be given by 
	\[(x,y)\cdot(x,y')=(x,xyy'+y+y')\]

	Then the thing here is the fiber (think vertical line in the plane!) over 0 is $\bbG_a$ and is isomorphic to $\bbG_m$ otherwise.
\end{ex}

\begin{ex}
	Let $\calE_\lambda=V(y^2z-x(x-z)(x-\lambda z))$ over $\Spec k[\lambda]$. Then when $\lambda=0$, we get the nodal cubic given by 
	$y^2z-x^2(x-z)$ (node at the origin).

	Now if you look at the connected component around 0 of $\Aut(\calE_\lambda)/\bbA_\lambda$, you actually find (when $\lambda=0$) 
	that $\bbG_m\cong\Aut(\calE_0)^0$.
\end{ex}

\subsection{Some definitions}
\begin{defn}
	A \textbf{homomorphism} $\phi:G\to G$ of group schemes over $S$ is a map $\phi:H\to G$ of schemes such that 
	\begin{center}
		\begin{tikzcd}
			H\times_S H\ar[r,"m_H"]\ar[d,"\phi\times\phi"] & H\ar[d]\ar[d,"\phi"]\\
			G\times_S G\ar[r,"m_G"] & G
		\end{tikzcd}
	\end{center}
\end{defn}
\begin{prob}
	Show that this automoatically imples that the identity and inversion maps are respected as well (automatically).
\end{prob}

\begin{defn}
	A \textbf{subgroup of $G\to S$} is a subscheme $H\subseteq G$ such that $H(T)\le G(T)$ for all $T$ over $S$.
\end{defn}
\begin{prob}
	Show that $\ker(\phi)\subseteq H$ is a subgroup.
\end{prob}
\begin{rmk}
	This gives you a nice way to construct new group schemes. For example, the following are exact:
	\[1\to \operatorname{SL}_n\to GL_n\xrightarrow{\det} \bbG_m\to 1\]
	and 
	\[1\to \bbG_m\to \GL_n\to \operatorname{PGL}_n\to 1\]
\end{rmk}

\begin{prop}
	Let $G\to S$ be a group scheme. Then $G\to S$ is separated if andy only if $e:S\to G$ is a closed immersion.
\end{prop}
\begin{prf}
	The idea here is that $S\to G$ is a closed immersion. Then we consider the map $m\circ(\id,S):G\times_SG\to G$
	and consider this along with the diagonal map $\Delta:G\to G\times_S G$ and this is a pullback square!
\end{prf}
\begin{cor}
	Any group scheme over $k$ is separated.
\end{cor}
The idea is going to be that if $X$ is any scheme over $k$, then any point $X\in X(k)$ is closed.

\end{document}
